### 2020-10-29 submission/manuscript.Rmd
Write Discussion

**The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure. First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs. Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations for each dataset to yield a significant test result. Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate. The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample. The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other. This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups []. When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B). However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S6B). Not including the small marine dataset, the average fraction the average fraction of falsely detecting a difference across datasets when only singletons were removed was `r 100*min_ave_p`%. If there is any relationship between the number of sequences and the treatment group, the risk of falsely rejecting the null hypothesis is inflated when researchers use the strategy of removing rare sequences. The most conservative approach is to not remove low abundance sequences.

@@

Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples. Such impacts will hinder the statistical power to differentiate between treatment groups. Instead of removing rare sequences, researchers should focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors. In addition, samples should be rarefied to a common number of sequences across samples without prior culling of rare sequences. The number of artifacts is correlated to the number of sequences being considered. With this in mind, rarefaction allows one to control for uneven sampling effort and to control for the number of artifacts in the analysis.

Need to treat rare sequences with a grain of salt

>>>>

**The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure. First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs. Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations for each dataset to yield a significant test result. Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate. The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample. The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other. This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [@Morris2013]. When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B). However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S6B). Not including the small marine dataset, the average fraction the average fraction of falsely detecting a difference across datasets when only singletons were removed was `r 100*min_ave_p`%. If there is any relationship between the number of sequences and the treatment group, the risk of falsely rejecting the null hypothesis is inflated when researchers use the strategy of removing rare sequences. The most conservative approach is to not remove low abundance sequences.

@@

### Big points
* Rare sequences are likely to be covered in other samples, suggesting they are rare; removing them biases against samples with fewer sequences
* Removing rare sequences reduces the diversity within samples and changes the shape of the communities
* The results include
  - increases the differences between samples; this makes it harder to detect differences between samples when those differences actually exist
  - If there is confounding between the number of reads per sample and the experimental grouping, then removing rare sequences increases the probability of falsely detecting a difference

Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples. Such impacts will hinder the statistical power to differentiate between treatment groups.

### What to do
* Removal of rare sequences is largely a response to optimizing the number of reads and length of sequences over their quality
* Researchers should instead focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors. We know that incomplete overlap of sequence reads significantly increases the sequencing error rate. Didn't test on crappy data
* Rarefy data to control for uneven sampling of communities and PCR or sequencing artifacts

In addition, samples should be rarefied to a common number of sequences across samples without prior culling of rare sequences. The number of artifacts is correlated to the number of sequences being considered. With this in mind, rarefaction allows one to control for uneven sampling effort and to control for the number of artifacts in the analysis.

### Should we implicitly trust rare sequences?
* No.
  - Need to treat rare sequences with a grain of salt, but don't discard
  - Need to use other methods to validate results observed using rare taxa
* For community metrics
  - rare sequencs are an important component and removing them shifts the community distribution.
  - metrics need to be treated on a relative rather than absolute basis

### Definition of rarity
* Idea of a blanket removal of reads below a threshold is stupid anyway
* Microbial communities contain numerically rare, temporarily rare, and spatially rare taxa.
* Removing those populations fundamentally changes how we think of microbial communities
